---
title: "R Notebook"
author: "Janet Roosimägi"
date: "18.12.2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Introduction

In my research I deal with modeling the stormwater drainage system in the town of Haapsalu. Most of the time I don´t need to analyze the data as much I have to process them. But there is also the data about the rain which is usually given in a second interval and makes the field of the data very large. This makes it difficult to analyse the data for a specific time period or day. Therefore, I set myself a goal to find an easier way to analyze the data by using R.

What I learned:

* reading data into R;
* viewing data;
* giving the names to the columns;
* reading data as character;
* making minute figures into hours;
* plotting
* creating a time series graph.


## Reading data into R:

```{r}
library(readxl)
Haapsalu_sademete_andmed_2014_2019_2_ <- read_excel("//intra.ttu.ee/home/janet.roosimagi/Desktop/EXX9067/Haapsalu sademete andmed/Haapsalu sademete andmed 2014-2019 (2).xlsx", 
                                                    col_types = c("date", "text"), skip = 1)
```

# View data:

```{r}
head(Haapsalu_sademete_andmed_2014_2019_2_,3)
```

## Giving a name to columns:

```{r}
names(Haapsalu_sademete_andmed_2014_2019_2_) <- c("Aeg",	"TunniSademeteSumma_mm")
```

## Reading data as character:

```{r}
Haapsalu_sademete_andmed_2014_2019_2_$Aeg_tekst<- as.character(Haapsalu_sademete_andmed_2014_2019_2_$Aeg)
```

## In order to make minute figures into hours, I "cut" minutes:

```{r}
substr(Haapsalu_sademete_andmed_2014_2019_2_$Aeg_tekst, 1, 10)
```

## Creating a new category (Aeg_paev):

```{r}
Haapsalu_sademete_andmed_2014_2019_2_$Aeg_paev<- as.factor(substr(Haapsalu_sademete_andmed_2014_2019_2_$Aeg_tekst, 1, 10))
```

## Finding the average values for the categories and plot the whole period and name the X and Y axis:

```{r}
with(Haapsalu_sademete_andmed_2014_2019_2_, tapply(Aeg_paev, TunniSademeteSumma_mm, mean))

plot(Haapsalu_sademete_andmed_2014_2019_2_$Aeg, Haapsalu_sademete_andmed_2014_2019_2_$`TunniSademeteSumma_mm`, xlab="Date", ylab="Sademed")
```


## Then I separate the years so I can make a graph for each year separately:

```{r}
Haapsalu_sademete_andmed_2014_2019_2_$aasta <- as.factor(substring(Haapsalu_sademete_andmed_2014_2019_2_$Aeg_tekst,1,4))

sademed2014 <- subset(Haapsalu_sademete_andmed_2014_2019_2_, aasta == '2014')

plot(sademed2014$Aeg, sademed2014$`TunniSademeteSumma_mm`)

sademed2015 <- subset(Haapsalu_sademete_andmed_2014_2019_2_, aasta == '2015')

plot(sademed2015$Aeg, sademed2015$`TunniSademeteSumma_mm`)

sademed2016 <- subset(Haapsalu_sademete_andmed_2014_2019_2_, aasta == '2016')

plot(sademed2016$Aeg, sademed2016$`TunniSademeteSumma_mm`)

sademed2017 <- subset(Haapsalu_sademete_andmed_2014_2019_2_, aasta == '2017')

plot(sademed2017$Aeg, sademed2017$`TunniSademeteSumma_mm`)

sademed2018 <- subset(Haapsalu_sademete_andmed_2014_2019_2_, aasta == '2018')

plot(sademed2018$Aeg, sademed2018$`TunniSademeteSumma_mm`)

sademed2019 <- subset(Haapsalu_sademete_andmed_2014_2019_2_, aasta == '2019')

plot(sademed2019$Aeg, sademed2019$`TunniSademeteSumma_mm`)
```


## Creating a time series graph:

```{r}
library(dygraphs)


Haapsalu_sademete_andmed_2014_2019_2_$datetime<- as.POSIXct(Haapsalu_sademete_andmed_2014_2019_2_$Aeg)

library(xts)

don <- xts(x = Haapsalu_sademete_andmed_2014_2019_2_$TunniSademeteSumma_mm, order.by = Haapsalu_sademete_andmed_2014_2019_2_$datetime)

dygraph(don) %>%
  dyOptions(labelsUTC = TRUE, fillGraph=TRUE, fillAlpha=0.1, drawGrid = FALSE, colors="#5ad6d8") %>%
  dyRangeSelector() %>%
  dyCrosshair(direction = "vertical") %>%
  dyHighlight(highlightCircleSize = 5, highlightSeriesBackgroundAlpha = 0.2, hideOnMouseOut = FALSE)  %>%
  dyRoller(rollPeriod = 1)

```

# Conclusion

In conclusion I can say that I find R very practical tool for analysing different data. In this course I got a good overview of R Studio and learned some useful tips how I can make my work easier by using R. What I am most happy about is the time series graph that I can use in my continuous work.


